Cost of Living in Toronto

### 2024-09-27 paper/paper.qmd
fixed cross references

The dataset published by `Social Development, Finance & Administration of Toronto` was last updated on August 2024 on an annual basis. However, the dataset uses price estimates from July 2022 - July 2023 for its income and expenses of the families. The variables chosen for this analysis were variables for the 11 households, for each entry of the income and expense types. See Table 1 for an overview. It is noted that expenses are not visible in `Table 1` due to the long list of income that occupies the first rows.

```{r}
#| label: low-income-families-preview
#| fig-cap: "Table 1: Sample of the Income/Expenses of Low Income Households in Toronto"
#| echo: false

@@

`Table 2` shows that income from employment was the major income source for all the 11 households in Toronto by a large margin, at approximately $1603 per month. This number also falls shy below the "low-income" threshold set by the City of Toronto, which is approximately $2208.5 [@citecityoftoronto].

```{r}
#| label: low-income-families-average-income
#| fig-cap: "Table 2: Average Income of Low Income Households in Toronto"
#| echo: false

@@

There is a huge price disparity between the expenses of the 11 families. The biggest expense was rent, at an average of $3146 per month, followed by food costs at $705. The huge cost rent highlights a need for further investigation and is discussed in @sec-discussion. See `Table 3` for an overview of the dataset. Note that all the values are rounded up and therefore may not add up to the total.

```{r}
#| label: low-income-families-average-expense
#| fig-cap: "Table 3: Average Expenses of Low Income Households in Toronto"
#| echo: false

@@

avg_expense %>% head(7) %>% knitr::kable()
```
## Difference between Income and Expense

To better understand the relationship between the income and expenses of the 11 families, they were merged together into a single dataset. In addition to this, a variable for their difference was also created. See `Table 4` for a preview of the dataset.

```{r}
#| label: low-income-families4
#| fig-cap: "Table 4: Differenece of Income and Expenses of Low Income Households in Toronto"
#| echo: false

@@

`Figure 1` shows income and expense for each household. Overall, expenses exceed income for all households. Only one household had an income that exceed expenses.

```{r}
#| label: low-income-families5
#| fig-cap: "Total Income and Expenses of Low Income Households in Toronto"
#| echo: false

@@

ggplot(data_long, aes(x = Household, y = Amount, fill = Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Figure 1: Total Income vs Total Expenses for Each Household", 
       x = "Household", 
       y = "Amount in $") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Furthermore, `Figure 2` gives us a better understanding of the spending behind the high expenses. Rent was by far the most expensive, comprising 66% of the expenses.

```{r}
#| label: low-income-families6
#| fig-cap: "Expense Breakdown of Low Income Households in Toronto"
#| echo: false

@@

Without context to this information, determining the efficiency of spending of the families is impossible. However, looking at the results suggest some inefficiencies in spending. For example, in `Figure 3`, the average rent was $3146 per month for the 11 families, a big difference from the average rent for a one-bedroom condo of $2,529 in Toronto for Q2 of 2023 [@citerentalmarket]. Expenses such as this could suggest that expenses could be correlated with poor financial decisions, however more research must be conducted to draw those conclusions.

>>>>

The dataset published by `Social Development, Finance & Administration of Toronto` was last updated on August 2024 on an annual basis. However, the dataset uses price estimates from July 2022 - July 2023 for its income and expenses of the families. The variables chosen for this analysis were variables for the 11 households, for each entry of the income and expense types. See @fig-preview for an overview. It is noted that expenses are not visible in @fig-preview due to the long list of income that occupies the first rows.

```{r}
#| label: fig-preview
#| fig-cap: "Sample of the Income/Expenses of Low Income Households in Toronto"
#| echo: false

@@

@fig-average-income shows that income from employment was the major income source for all the 11 households in Toronto by a large margin, at approximately $1603 per month. This number also falls shy below the "low-income" threshold set by the City of Toronto, which is approximately $2208.5 [@citecityoftoronto].

```{r}
#| label: fig-average-income
#| fig-cap: "Average Income of Low Income Households in Toronto"
#| echo: false

@@

There is a huge price disparity between the expenses of the 11 families. The biggest expense was rent, at an average of $3146 per month, followed by food costs at $705. The huge cost rent highlights a need for further investigation and is discussed in @sec-discussion. See @fig-average-expense for an overview of the dataset. Note that all the values are rounded up and therefore may not add up to the total.

```{r}
#| label: fig-average-expense
#| fig-cap: "Average Expenses of Low Income Households in Toronto"
#| echo: false

@@

avg_expense %>% head(7) %>% knitr::kable()
```

## Difference between Income and Expense

To better understand the relationship between the income and expenses of the 11 families, they were merged together into a single dataset. In addition to this, a variable for their difference was also created. See @fig-income-expense for a preview of the dataset.

```{r}
#| label: fig-income-expense
#| fig-cap: "Differenece of Income and Expenses of Low Income Households in Toronto"
#| echo: false

@@

@fig-difference shows income and expense for each household. Overall, expenses exceed income for all households. Only one household had an income that exceed expenses.

```{r}
#| label: fig-difference
#| fig-cap: "Total Income and Expenses of Low Income Households in Toronto"
#| echo: false

@@

ggplot(data_long, aes(x = Household, y = Amount, fill = Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Figure 1: Total Income vs Total Expenses for Each Household", 
       x = "Household", 
       y = "Amount in $") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Furthermore, @fig-breakdown gives us a better understanding of the spending behind the high expenses. Rent was by far the most expensive, comprising 66% of the expenses.

```{r}
#| label: fig-breakdown
#| fig-cap: "Expense Breakdown of Low Income Households in Toronto"
#| echo: false

@@

Without context to this information, determining the efficiency of spending of the families is impossible. However, looking at the results suggest some inefficiencies in spending. For example, in @fig-average-expense, the average rent was $3146 per month for the 11 families, a big difference from the average rent for a one-bedroom condo of $2,529 in Toronto for Q2 of 2023 [@citerentalmarket]. Expenses such as this could suggest that expenses could be correlated with poor financial decisions, however more research must be conducted to draw those conclusions.
